---
title: "downstreamGWAS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{downstreamGWAS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
library(downstreamGWAS)
```

# Introduction

`downstreamGWAS` is an R package that utilizes that standardized format of `tidyGWAS().`

With column names and data-types preconfigured, many downstream analysis is just a step of inputting file paths into the desired software (the summary statistics, along with any potential reference data), as well as a set of parameters.

The idea behind `downstreamGWAS` is to put this all together in one place, and to automate the oftentimes laborous steps of running a downstream analysis. Through containerization of external software, there is no additional dependency except apptainer/singularity installed on the computing system.

Jobs are constructed within R, and can then be written out to disk as a slurm job, or programatically edited within the R environment. `downstreamGWAS` makes it extremely easy to run downstream analysis, and to scale analysis to hundreds of summary statistics.

```{r}

```

In it's essence, downstreamGWAS is just a template package, abstracting away the steps of matching filepaths. In addition, the templating of containers allows us to abstract away any potential software installation problems.
