---
title: "Colocalisation analysis with run_coloc()"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Overview

Colocalisation tests whether two traits share a causal variant at a genomic
locus. downstreamGWAS wraps `coloc::coloc.abf()` with `run_coloc()`, which
takes two tidyGWAS directories and a genomic region and handles the data
extraction, formatting, and trait-type detection automatically.

The typical workflow is:

1. **Clump** one trait to identify genome-wide significant loci
2. **Run coloc** at each locus, testing against a second trait

```{r setup}
library(downstreamGWAS)
```

# Step 1: Identify loci by clumping

Before running colocalisation, you need regions to test. `pipeline_clumping()`
produces a `merged_loci.bed` file — each row is a merged LD-independent locus
with genomic coordinates.

```{r}
pipeline_clumping(
  parent_dir = "/path/to/trait1",
  execute = TRUE,
  prepare_inputs = TRUE,
  schedule = schedule_slurm(account = "my-project", mem = "8gb")
)

# Read the resulting loci
loci <- readr::read_tsv(
  "/path/to/trait1/analysis/clumping/merged_loci.bed",
  col_names = c("chr", "start", "end", "n_snps", "pvals", "lead_snps")
)
loci
```

# Step 2: Run coloc at a single locus

`run_coloc()` takes two tidyGWAS directories, a chromosome, and start/end
coordinates. It extracts the variants in the region from both datasets,
intersects them, and runs `coloc::coloc.abf()`.

```{r}
result <- run_coloc(
  parent_dir  = "/path/to/trait1",
  parent_dir2 = "/path/to/trait2",
  chrom = "7",
  start = 1788081,
  end   = 2289862
)

result$summary
```

The output is the standard `coloc.abf()` return object. The key value is
`PP.H4.abf` — the posterior probability that both traits share a single causal
variant.

## Formatting the result

`format_coloc()` extracts the key numbers into a tidy tibble:

```{r}
format_coloc(result, name = "trait1_vs_trait2_chr7")
```

# Step 3: Loop over all loci

To test all clumped loci against a second trait:

```{r}
results <- purrr::pmap(loci, function(chr, start, end, ...) {
  run_coloc(
    parent_dir  = "/path/to/trait1",
    parent_dir2 = "/path/to/trait2",
    chrom = chr,
    start = start,
    end   = end
  )
})

# Remove loci where trait2 had no signal (returns NULL)
results <- purrr::compact(results)

# Format into a single table
purrr::imap_dfr(results, function(res, i) {
  format_coloc(res, name = paste0("locus_", i))
})
```

# Parameters and when to change them

## `min_pval` (default: `5e-08`)

Controls the minimum p-value required in **trait 2** at the locus before
running coloc. If no variant in trait 2 reaches this threshold, `run_coloc()`
returns `NULL` early. This saves compute time by skipping loci where the second
trait has no signal at all.

**When to change it:** Relax to `1e-05` or `1e-04` if you want to test
colocalisation even at loci where trait 2 has only suggestive significance.
This is common when testing against eQTL datasets, where sample sizes may be
smaller. Be aware that coloc results at loci without clear association in both
traits are harder to interpret.

```{r}
# More permissive — test even with weak signal
run_coloc(..., min_pval = 1e-4)
```

## `trait_type1` / `trait_type2` (default: `"guess"`)

Tells coloc whether each trait is case-control (`"cc"`) or quantitative
(`"quant"`). By default, `run_coloc()` guesses by checking whether `CaseN`
exists in the dataset — if it does, it assumes case-control.

**When to change it:** Override the guess when you know it is wrong. For
example, a dataset might carry `CaseN` from its original source but you want
to treat it as quantitative for the purpose of coloc. For case-control traits,
`EffectiveN` is used as `N` when available.

```{r}
# Explicitly set trait types
run_coloc(..., trait_type1 = "cc", trait_type2 = "quant")
```

## Priors: `p1`, `p2`, `p12`

These are the prior probabilities that a SNP is associated with trait 1 only
(`p1`), trait 2 only (`p2`), or both traits (`p12`). The defaults
(`p1 = 1e-4`, `p2 = 1e-4`, `p12 = 1e-5`) are from the coloc package and are
reasonable for most GWAS-vs-GWAS comparisons.

**When to change them:** The ratio `p12 / p1` (equivalently `p12 / p2`)
encodes your prior belief about how likely two traits are to share a causal
variant, given that at least one is associated. The default ratio is 0.1 —
i.e., if a SNP is causal for trait 1, there is a 10% prior chance it is also
causal for trait 2.

- For traits you expect to share biology (e.g. two psychiatric disorders),
  increase `p12` toward `1e-4` (ratio ~1).
- For traits with little expected overlap (e.g. a metabolite vs. a behavioural
  trait), the defaults are fine or you could decrease `p12` to `1e-6`.
- `coloc::sensitivity()` can be used to check how sensitive your results are
  to the choice of priors.

```{r}
# Stronger prior for shared causality
result <- run_coloc(..., p12 = 5e-5)

# Check sensitivity to prior choice
coloc::sensitivity(result, rule = "H4 > 0.8")
```

# Missing EAF

coloc requires allele frequencies. If `EAF` is missing from a dataset,
`run_coloc()` attempts to impute it from a 1000 Genomes EUR reference panel
(looked up via the `dbsnp` environment variable). This works for common
variants but may not cover all SNPs. If EAF cannot be resolved, those variants
are dropped from the analysis — check the log output for how many variants
remain after filtering.

# Interpreting results

| Hypothesis | Meaning |
|-----------|---------|
| H0 | No association with either trait |
| H1 | Association with trait 1 only |
| H2 | Association with trait 2 only |
| H3 | Both traits associated, but different causal variants |
| H4 | Both traits share a single causal variant |

A high `PP.H4` (e.g. > 0.8) provides evidence for colocalisation. A high
`PP.H3` suggests both traits have signal at the locus but driven by different
variants — this distinction is important and often confused with
colocalisation in naive overlap analyses.
